### 2016-11-11 Manuscript/SW_early_life_telomeres.Rmd
Home eds

#####Telomere dynamics and age
Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM: estimate $\pm$ s.e. = `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,1])` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,2])`; t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID/LayYear))$tTable[2,5])`). We found that modelling a non-linear effect age explained almost double the variation in RTL (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than a linear model `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`, suggesting that telomere length decreased at the greatest rate in early life. Accordingly, RTL decreased with age at a greater rate in juveniles (estimate $\pm$ s.e. = `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,2])`) compared to adults (estimate $\pm$ s.e. = `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,2])`). Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig. 1A).

@@

#####Spatiotemporal variation in telomere dynamics and the environment
There was significant variation in RTL among cohorts (), although we found no evidence that the relationshio between telomere length and age varied among cohorts. Rather, cohort-level differences in RTL were consistent over the lifespan of Seychelles warblers.

A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1). Model selection revealed that sex only appeared in models where tarsus length was also present (Table S1), and sex was no longer significant when tarsus length was removed from the full model (P = ). RTL increased with tarsus length in males, but decreased in females. In general, the model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-average confidence intervals not overlapping zero (Table S1; Fig. S1).

>>>>

#####Telomere dynamics and age
Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM with individual ID as random effect: estimate $\pm$ s.e. = `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,1])` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,2])`; t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`). We found that modelling a non-linear effect age explained almost double the variation in RTL (R^2^ of linear regression of telomere length and log age = `r round3(summary(lm(RTL~log(Agemonths),data=dd))$r.squared)` than fitting a linear relationship `r round3(summary(lm(RTL~Agemonths,data=dd))$r.squared)`, suggesting that telomere length decreased at the greatest rate in early life. Accordingly, RTL decreased with age at a greater rate in juveniles (estimate $\pm$ s.e. = `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID/LayYear))$tTable[2,2])`) compared to adults (estimate $\pm$ s.e. = `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,1],lessthan = F)` $\pm$ `r round3(summary(lme(RTL~Agemonths,data=subset(dd,Ageclass == 'A'),random=~1|BirdID/LayYear))$tTable[2,2])`). Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig. 1).

Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (), and was a significantly better model than one including catch year as a random effect. This is indicative of cohort effects on telomere length across the Seychelles warbler lifespan. Furthermore, we found no evidence that the relationship between telomere length and age varied among cohorts (). Rather, cohort-level differences in RTL were consistent over the lifespan of Seychelles warblers.

@@

#####Spatiotemporal variation in telomere dynamics and the environment
A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1). Model selection revealed that sex only appeared in models where tarsus length was also present (Table S1), and sex was no longer significant when tarsus length was removed from the full model (P = ). RTL increased with tarsus length in males, but decreased in females. In general, the model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-average confidence intervals not overlapping zero (Table S1; Fig. S1).
